OLS Regression Project

### 2021-01-04 resources/paper_calenviro.Rmd
finished paper, except for formatting (how do I get it to output a PDF?)

---
output:
  pdf_document:
    citation_package: natbib
    fig_caption: yes
    template: ./resources/aog-latex-ms.tex
  word_document:
    reference_docx: ./resources/aog_word_style.docx
fontfamily: mathpazo
fontsize: 11pt
anonymous: false
endnotes: false
pagenumber: 1
#bibliography: ../project.bib
#biblio-style: ./resources/ajs.bst
title: "Low Birth-Weight and Air Pollution: Examining Some Associations in CalEnviroScreen 3.0 Data"
author:
- affiliation: University of Oregon, Sociology
  name: "David Purucker""
keywords: "environmental justice, low birth weight, poverty, cumulative pollution burden, public health"
thanks: "My eternal gratitude to Natasha Erickson."
abstract: "How do environmental pollutants and socioeconomic factors correlate with the incidence of low birth-weight?"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
source("check_packages.R")
source("useful_functions.R")
```

@@

To study this question, I used CalEnviroScreen 3.0, a public health modeling tool designed to assess the 'cumulative impacts' of environmental and social stressors on Californians (Rodriquez and Zeise 2017). Cumulative impact refers to the totality of health risks affecting a population, including both environmental and socio-economic conditions. In this analysis, OLS regrssion models are used to analyze the correlations of various pollution measures from CalEnviroScreen with the incidence of LBW at the census tract level, to determine the prevalence of environmental and social correlates of LBW in the CES data, and understand the strengh of different LBW effects of each stressor. Findings indicate a positive and moderately strong independent effect of ozone air pollution on LBW, and similarly positive, though weaker, effects from particulate matter air pollution. These findings suggest that policies aimed at curtailing air pollution, especially ozone, will have the most beneficial effect on reducing LBW and its associated public health effects. [pending discussion section writeup]

@@

The connection between LBW and other forms of pollution included in the CalEnviroScreen data is also well-established. LBW correlates have been found in studies considering proximity to freeway traffic (Laurent et al. 2013, Ponce et al. 2005), pesticides (Harley et al. 2011), diesel particulate matter (Laurent et al. 2016), contaminated drinking water (Currie et al. 2013) and proximity to hazardous waste sites (Berry and Bove 1997). Finally, social determinants of health such as porverty and race also have a well-established relationship with LBW (Lu and Hafton 2003, Martinson and Recihman 2016). [expand—what do Lu and Hafton and Martinson and Reciman find][state what my study adds hahaha][relatively new dataset, high-quality census tract data over a large geographic area]

@@

``` {r descriptive statistics—-LBW distribution}

@@

``` {r descriptive statistics--ozone and LBW; run chunk together}
# Joint distribution of ozone and LBW
plot(ces3_complete$ozone, ces3_complete$lbw, main="Fig 2. Census Tracts by Ozone and Low Birth-Weight", xlab="Ozone", ylab ="Low Birth-Weight %")
abline(lm(ces3_complete$lbw~ces3_complete$ozone))

@@

Four regression models were constructed for this analysis. Note that because the range of ozone data only ranges from 0.02 to 0.068 parts per million, it is more meaningful to interpret the coefficients for this variable by considering the effect of a 1/100th-unit increase, rather than a 1-unit increase. Therefore, in the following interrpretations, I divide the ozone coefficient by 100.

@@

Model 3 assesses the independent effect of race on LBW. This model indicates some surprising findings. It predicts that for every 1 percentage point increase in the nonwhite population in census tracts, there will be a .02 percentage point increase in LBW. The race variable here is statistically significant at the .001 level. Inclusion of race in this model changes some predicted associations of environmental pollutants and LBW present in the earlier models. The effect of ozone actually increases with the inclusion of race, from .17 units in model 1 to .20 here. [ask Aaron] The effect of diesel PM pollution remains 0.01. PM 2.5 pollution, however, is no longer statistically significant in this model, and the magnitude of its effect has dropped to almost 0, from .05 in model 1 to -0.01. PM 2.5's statistical insignificance in this model suggests that a higher nonwhite population in census tracts accounts for the association of PM2.5 pollution with LBW in models 1 and 2. Similar to model 1, ozone and diesel PM remain stasticially significant. Some of the other environmental pollutants——drinking water, pesticides, and toxic releases——are statistically significant in this model, but their magnitudes remain small.

@@

``` {r Models}

@@

stargazer(model1, model2, model3, model4, type="text", digits = 3, covariate.labels = c("Ozone", "PM2.5", "Diesel PM", "Drinking Water Quality", "Pesticides", "Toxic Releases", "Traffic", "Proximity to Cleanup Sites", "Groundwater", "Hazardous Wastes", "Impaired Water", "Solid Waste", "Poverty", "Nonwhite Tract Percentage"), dep.var.labels = "Low Birth-Weight")

@@

¶ resume

¶ implications for the study of low birth-weight and public health

¶ future directions

>>>>

---
title: "OLS Regression Project"
author: "David Purucker"
output:
  word_document: default
  pdf_document: default
  html_document:
    df_print: paged
---

```{r Setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, error=FALSE, message=FALSE, include=FALSE)
load("~/Desktop/sociology/soc_year2/Soc 412:512 (Stats II)/calenviro_proj/output/ces3_complete.RData")
if(!require(stargazer)) {
  install.packages("stargazer", "texreg")
}
```

@@

To study this question, I used CalEnviroScreen 3.0, a public health modeling tool designed to assess the 'cumulative impacts' of environmental and social stressors on Californians (Rodriquez and Zeise 2017). Cumulative impact refers to the totality of health risks affecting a population, including both environmental and socio-economic conditions. In this analysis, OLS regrssion models are used to analyze the correlations of various pollution measures from CalEnviroScreen with the incidence of LBW at the census tract level, to determine the prevalence of environmental and social correlates of LBW in the CES data, and understand the strengh of different LBW effects of each stressor. Findings indicate a positive and moderately strong independent effect of ozone air pollution on LBW, and similarly positive, though weaker, effects from particulate matter air pollution. These findings suggest that policies aimed at curtailing air pollution, especially ozone, will have the most beneficial effect on reducing LBW and its associated public health effects.

@@

The connection between LBW and other forms of pollution included in the CalEnviroScreen data is also well-established. LBW correlates have been found in studies considering proximity to freeway traffic (Laurent et al. 2013, Ponce et al. 2005), pesticides (Harley et al. 2011), diesel particulate matter (Laurent et al. 2016), contaminated drinking water (Currie et al. 2013) and proximity to hazardous waste sites (Berry and Bove 1997). Finally, social determinants of health such as porverty and race also have a well-established relationship with LBW (Lu and Hafton 2003, Martinson and Recihman 2016).

@@

``` {r descriptive statistics—-LBW distribution, echo=FALSE, message=FALSE, warning=FALSE, results="asis"}

@@

``` {r descriptive statistics--ozone and LBW, echo=FALSE, message=FALSE, warning=FALSE, results="asis"}
# Joint distribution of ozone and LBW
plot(ces3_complete$ozone, ces3_complete$lbw, main="Fig 2. Census Tracts by Ozone and Low Birth-Weight", xlab="Ozone", ylab ="Low Birth-Weight %")
abline(lm(ces3_complete$lbw~ces3_complete$ozone))

@@

Four regression models were constructed for this analysis. Note that because the ozone data only ranges from 0.02 to 0.068 parts per million, it is more meaningful to interpret the coefficients for this variable by considering the effect of a 1/100th-unit increase, rather than a 1-unit increase. Therefore, in the following interrpretations, I divide the ozone coefficient by 100.

@@

Model 3 assesses the independent effect of race on LBW. This model indicates some surprising findings. It predicts that for every 1 percentage point increase in the nonwhite population in census tracts, there will be a .02 percentage point increase in LBW. The race variable in this model is statistically significant at the .001 level. The inclusion of race changes some predicted associations of environmental pollutants and LBW present in the earlier models. The effect of ozone actually increases with the inclusion of race, from .17 units in model 1 to .20 here [ask Aaron about this]. The effect of diesel PM pollution remains 0.01. PM 2.5 pollution, however, is no longer statistically significant in this model, and the magnitude of its effect has dropped to almost 0, from .05 in model 1 to -0.01. PM 2.5's statistical insignificance in this model suggests that a higher nonwhite population in census tracts accounts for the association of PM2.5 pollution with LBW in models 1 and 2. Similar to model 1, ozone and diesel PM remain stasticially significant. Some of the other environmental pollutants——drinking water, pesticides, and toxic releases——are statistically significant in this model, but their magnitudes remain small.

@@

``` {r Models, echo=FALSE, message=FALSE, warning=FALSE, results="asis"}

@@

library(stargazer)
stargazer(model1, model2, model3, model4, type="text", digits = 3, covariate.labels = c("Ozone", "PM2.5", "Diesel PM", "Drinking Water Quality", "Pesticides", "Toxic Releases", "Traffic", "Proximity to Cleanup Sites", "Groundwater", "Hazardous Wastes", "Impaired Water", "Solid Waste", "Poverty", "Nonwhite Tract Percentage"), dep.var.labels = "Low Birth-Weight")

@@

In this analysis, I have applied an OLS regression method to examine the influence of various environmental and social stressors upon low birth-weight (LBW) outcomes in the CalEnviroScreen 3.0 data analysis tool. Four models were constructed to assess the relative strengh of correlation between low birth-weight and variables like ozone, particulate matter air pollution, hazardous wastes, and social conditions like poverty or race. From this analysis, three findings seem particularly important. First, air pollution, including ozone, PM2.5 pollution, and diesel PM, has a stronger effect on LBW outcomes in California census tracts than other forms of pollution, such as proximity to hazardous waste sites, impaired water quality, or pesticide exposure. While this analysis doesn't allow me to determine the precise causal mechanism behind the different strength of association for these variables, it seems reasonable to assume that air pollution has a stronger effect on many kind of health outcomes, including LBW, because of its very wide geographical dispersion. Other forms of pollution, such as exposure to pesticides or hazardous wastes, have a strong demonstrated effect upon adverse birth outcomse at the local level, but do not pose nearly as great a risk over a wide geographic area.

The models which include poverty and race controls reveal two other interesting findings. First, when race is controlled for (in model 3), PM2.5 is no longer statistically significant, and the magnitude of its effect drops significantly. This indicates that a higher nonwhite population in census tracts accounts for the association of PM2.5 pollution with LBW present in the model which does not include any race variable. Second, the model which controls for both poverty and race indicates that while the inclusion of these variables reduces the magnitude and statistical significance of PM2.5, the other air pollution effects, ozone and diesel PM, retain a significant independent effect on LBW. This is especially true of ozone. The role of ozone is particularly interesting in this analysis. In model 2, when poverty is controlled for, the effect of ozone diminishes considerably. In model 3, however, when race is controlled for, the effect of ozone actually increases from its level in model 1. Finally, in model 4, its effect returns almost exactly to its level in model 1. It is not clear to me why this is the case. Future research could explore the particular relationship of ozone in the presence or absence of different, or more disaggregated, socio-economic variables of this kind.

The main implication of this study seems to be that, among all major environmental stressors, the most significant threats to public health are air pollution, especially ozone. Policies aimed at reducing adverse birth outcomes, which have a disproportionate effect upon non-white populations, should prioritize efforts to reduce these pervasive forms of pollution in California and elsewhere. Doing so would have a particularly positive effect for poor and non-white people. However, this analysis also indicates that there is an independent effect of poverty and racialization above and beyond the effects of air pollution. Policies aimed at reducing poor birth outcomes will have to take this into account.

This analysis has a number of important limitations. First, this data only includes census tracts in California. While results in a state as large and socially diverse as California are plausibly generalizable to many other contexts, future research on pollution and LBW should explore the relationship in other contexts and at different scales. Second, future analyses could use more complex methods to explore interactions between different variables. Third, more detailed socio-economic data than is available in CalEnviroScreen could help to explain puzzles revealed in this analysis, such as the up-and-down strength of the ozone effect on LBW.
